\\mytitle

### 2017-06-10 pdocs/thesis/epj-jcgt.tex
jcgt article started

% !TeX root = article.tex
\section{Description of plasticity in the framework of physics engines}

In this section, key concepts related to the introduced model are explained. The main differences between 
traditional structural analysis and physics engines based approaches are reviewed and discussed.

Velocity-based formulation of constraint based rigid body simulation 
is  commonly used by physics based game
developers and film production teams.
%\citet[p.~45]{erleben.thesis} 
\cite{erleben.thesis} 
provides reasoning and theoretical details for the popularity of 
velocity-based formulation in constraint-based rigid body simulation instead of accelaration based. 
The main reason is that collision handling can be done without the use of additional procedures.

Work presented by  
\cite{erleben.thesis} provides the basis for the velocity-based formulation discussed in this work.
%\citet[p.~45-50]{erleben.thesis}.
% pdf page 64
In the following section, these formulations will be clarified by a simple example using \cbullet\ implementation.

Impulse $\vec{J}$
in the time interval $\Delta t $ can be written as:
\begin{equation} \label{eq:impulseIntegral}
\vec{J} = \int_{0}^{\Delta t} \vec{f}_{true}(t) dt,
\end{equation}
where $\vec{f}_{true}(t)$ is force.

Using Newton's second law of motion $\vec{F}=m\vec{a}$ ,
$\vec{v}^{\Delta t}$ can be solved for the velocity as:
\begin{equation} \label{eq:impulseIntegraWithNewton}
\int_{0}^{\Delta t} m \frac{d\vec{v}}{dt}dt= \int_{0}^{\Delta t} \vec{f}_{true}(t)
\end{equation}
\begin{equation} \label{eq:impulse}
m(\vec{v}^{\, \Delta t} - \vec{v}^{\, 0})=\vec{J},
\end{equation}
where superscripts denote time, i.e. ${\vec{v}}^{\Delta t}=\vec{v}(\Delta t)$.
Next position can be found
by integrating the velocity.
Updates after each step can be summarized  for locations and  
for velocities respectively as follows:

\begin{equation} \label{eq:eomL} % pdf page 69
\vec{s}^{\, t+\Delta t} = \vec{s}^{\, t}+\Delta t S \vec{u}^{\, t+\Delta t}
\end{equation}

\begin{equation} \label{eq:eomV}
\vec{u}^{\, t+\Delta t} = \vec{u}^{\, t}+\Delta t M^{-1}(C N \vec{f}^{\ t+\Delta t} + \vec{f}_{ext}) .
\end{equation}

The symbols used in Equations \ref{eq:eomL} and \ref{eq:eomV}
are summarized in Table \ref{tab:eom}.
Figure \ref{fig:eom-contact} describes the collision of two
bodies, rectangular body $B_1$ and triangular body $B_2$
where $\vec{r}_i$ is position of body $i$,
$\vec{p}_k $ is position of contact point $k$,
$\vec{r}_{ik} $ is a vector between the center of gravity of body $i$ and contact point $k$, 
and
$\vec{n}_{k}$ is the contact normal for contact point $k$.
It is common convention that the contact normal points
from the body with the smallest index to the body with the largest index, \cite{erleben.thesis}.
In case of point or edge contacts, averaging the normals of neighboring polygons can be used, \cite{Hahn:1998}.

\begin{figure}[tb!]
\centering
\begin{tikzpicture}
\coordinate (O1) at(2,1);
\coordinate (O2) at(2,4);
\coordinate (C) at(2,2);
\draw (0,0) -- (4,0) -- (4,2) -- (0,2) --(0,0);
\draw (2,2) -- (3.5,5) -- (0.5,5) -- (2,2) ;
\node at (3.5,0.4) {$B_1$};
\filldraw (O1) circle (0.5mm) node[anchor=north] {$\vec{r}_1$};
\node at (2.8,4.5) {$B_2$};
\filldraw (O2) circle (0.5mm) node[anchor=south] {$\vec{r}_2$};
\filldraw (C) circle (0.5mm) node[anchor=north west] {$\vec{p}_1$};
\draw[-{Stealth[length=3mm]}] (O1) -- (C) node[anchor=north east] {$\vec{r}_{11}$};
\draw[-{Stealth[length=3mm]}] (O2) -- (C) node[anchor=south east] {$\vec{r}_{21}$};
\draw[-latex,thick] (C) -- ++(0,1.4) node[anchor=west] {$\vec{n}_{1}$};
\node[anchor=west] at (4.5,4.5) {
$\vec{r}_i$  =  position of body $i$
};
\node[anchor=west] at (4.5,4) {$\vec{p}_k $ = position of contact point $k$};
\node[anchor=west] at (4.5,3.5) {$\vec{r}_{ik} $ = $\vec{p}_k - \vec{r}_i $};
\node[anchor=west] at (4.5,3) {$\vec{n}_{k} $ = normal for contact point $k$};
\end{tikzpicture}
\caption{Illustration of nomenclature for equations of motion for collision.}
\label{fig:eom-contact}
\end{figure}

% pdf page 33, notation in typical ODEs
\begin{table}
\tbl{Nomenclature for equations of motion}{
\begin{tabular}{|l| l|}
\hline
{\bf Symbol} & {\bf Description} \\  \hline
$\vec{r}_i$ & position of center of mass for body $i$  \\ \hline
$\vec{q}_i$ & orientation for body $i$ as quaternion $\lbrack s_i, x_i, y_i, z_i \rbrack ^T $ \\
\hline
$\vec{p}_h$ & contact or joint point $k$  \\ \hline
$\vec{r}_{ki}$ & $\vec{p}_k - \vec{r}_i$  \\ \hline
$\vec{s}$ & $\lbrack \vec{r}_1, \vec{q}_1,...,\vec{r}_n, \vec{q}_n \rbrack ^T $\\ \hline
$Q_i$ & \begin{tabular}{@{}c}
rotation of quaternion $\vec{q}_i$
as matrix \\ where 
$\frac{1}{2}\vec{\omega}_i \vec{q}_i=Q_i \vec{\omega}_i$
\end{tabular}
   $
\frac{1}{2} \left[ \begin{array}{ccc}
-x_i & -y_i & -z_i \\
s_i & z_i & -y_i \\
-z_i & s_i & x_i \\
y_i & -x_i & s_i
\end{array} \right]
$
\\ \hline
$S$ & 
\begin{tabular}{@{}c}
generalized transformation matrix \\
$ S \in \mathbb{R}^{7n \times 6n}$
\end{tabular}
 $ \left[ \begin{array}{ccccc}
1 &  &  & & 0 \\
 & Q_i  \\
 & & \ddots  \\
 & & & 1 \\
0 & & & & Q_n 
\end{array} \right]
$
\\ \hline
$\vec{v}_i$ & linear velocity of  center of mass for body $i$   \\ \hline
$\vec{\omega}_i$ & angular velocity of center of mass for body $i$  \\ \hline
$\vec{u}$ & $\lbrack \vec{v}_1, \vec{\omega}_1,...,\vec{v}_n, \vec{\omega}_n \rbrack ^T $\\ \hline
$M$ &
\begin{tabular}{@{}c}
 generalized mass matrix \\
$ M \in \mathbb{R}^{6n \times 6n}$  
\end{tabular}
$
\left[ \begin{array}{ccccc}
m_i 1 &  &  & & 0 \\
 & I_1  \\
 & & \ddots  \\
 & & & m_n 1 \\
0 & & & & I_n 
\end{array} \right]
$
\\ \hline
$I_i$ & inertia tensor for body $i$  \\ \hline
$C$ & contact condition matrix  $ C \in \mathbb{R}^{6n \times 3K}$ \\ \hline
$N$ & contact normal matrix  $ N \in \mathbb{R}^{3K \times K}$ \\ \hline
\end {tabular}}
\label{tab:eom}
\end{table}

Friction in contacts and joint constraints can be handled in a unified way by refactoring
equation \ref{eq:eomV} as,  
\cite{erleben.thesis}
%\citet[p.~66-67]{erleben.thesis}
\begin{equation} \label{eq:eomV2}
\vec{u}^{\, t+\Delta t} = \vec{u}^{\, t}+\Delta t M^{-1}(
 J_{contact}^T \vec{\lambda}_{contact}
+ J_{joint}^T \vec{\lambda}_{joint}
+ \vec{f}_{ext}),
\end{equation}
where Jacobian terms $J_{joint}^T$ for joints are 
derived by taking time derivatives of the kinematic constraints.
Symbols used in Equation \ref{eq:eomV2} are summarized in Table
\ref{tab:eom-g} and Figure \ref{fig:eom-joint},
where $\vec{r}_{anc}^{\,i}$ is used to define at which point
joint constraint is applied relative to body $i$.

\begin{figure}
\centering
\begin{tikzpicture}
\coordinate (O1) at(1,1);
\coordinate (O2) at(1,3);
\coordinate (C) at(1,2);
\draw (0,0) -- (2,0) -- (2,2) -- (0,2) --(0,0);
\draw (0,2) -- (2,2) -- (2,4) -- (0,4) --(0,2) ;
\node at (0.5,0.4) {$B_1$};
\filldraw (O1) circle (0.5mm) node[anchor=north] {$\vec{r}_1$};
\node at (0.5,3.5) {$B_2$};
\filldraw (O2) circle (0.5mm) node[anchor=south] {$\vec{r}_2$};
\draw[-{Stealth[length=3mm]}] (O1) -- (C) node[anchor=north east] {$\vec{r}_{anc}^{\,1}$};
\draw[-{Stealth[length=3mm]}] (O2) -- (C) node[anchor=south east] {$\vec{r}_{anc}^{\,2}$};
\node[anchor=west] at (4.5,3.5) {
$\vec{r}_i$  =  position of body $i$ 
};
\node[anchor=west] at (4.5,3) {$\vec{r}_{anc}^{\,i} $ = body frame vector $i$};
\end{tikzpicture}
\caption{Illustration of nomenclature for equations of motion for joint.}
\label{fig:eom-joint}
\end{figure}

% pdf page 33, notation in typical ODEs
\begin{table}
\tbl{Additional terms  for generalized equations of motion}{
\begin{tabular}{|l| l|}
\hline
{\bf Symbol} & {\bf Description} \\  \hline
$J_{contact}$ & Jacobian matrix for contacts  \\ \hline
$\lambda_{contact}$ & vector of lagrange multipliers for contacts  \\ \hline
$J_{joint}$ & Jacobian matrix for joints  \\ \hline
$\lambda_{joint}$ & vector of lagrange multipliers for joints  \\ \hline
\end {tabular}}
\label{tab:eom-g}
\end {table}

Constraint processing in \cbullet\ is based on ODE, \cite{ode}.
Joints are also discussed in detail in  
\cite{erleben.thesis}.
%\citet[p.~60-90]{erleben.thesis}.
Equations \ref{eq:constraintEquation}, \ref{eq:lambdaLow} and
\ref{eq:lambdaHigh} 
are created for each constraint. 
Derivation for terms in Equation \ref{eq:constraintEquation}
can be done using the position and orientation of connected bodies
e.g. for ball joint formulation is based on both joint points having the same position.
In contact cases, formulation is easier if it is done using velocities, \cite{ode.joints}.

\begin{equation} \label{eq:constraintEquation}
J_1 \vec{v}_1 + \Omega_1 \vec{\omega}_1 + 
J_2 \vec{v}_2 + \Omega_2 \vec{\omega}_2 = \vec{c} + C \vec{\lambda}
\end{equation}

\begin{equation} \label{eq:lambdaLow}
\vec{\lambda} \geq \vec{l}
\end{equation}

\begin{equation} \label{eq:lambdaHigh}
\vec{\lambda} \leq \vec{h}
\end{equation}

In the following section, these equations will be explained by a simple example.
The main parameters  and corresponding fields in \cbullet\  
are given in Table \ref{tab:constraintParameters}.

\begin {table}
\tbl {Constraint parameters}{
\begin{tabular}{|c| l| l|}
\hline
{\bf Parameter} & {\bf Description} & {\bf btConstraintInfo2 pointer}\\  \hline
$J_1, \Omega_1$ & Jacobian & m\_J1linearAxis, m\_J1angularAxis \\
$J_2, \Omega_2$ & & m\_J2linearAxis, m\_J2angularAxis \\ \hline
$\vec{v}$ & linear velocity & \\ \hline
$\vec{\omega}$ & angular velocity & \\ \hline
$\vec{c}$        &  right side vector   & m\_constraintError \\ \hline
$C$  & constraint force mixing & cfm \\  \hline
$\vec{\lambda}$ & constraint force &  \\ \hline
$\vec{l}$ & low limit for constraint force & m\_lowerLimit \\ \hline
$\vec{h}$ & high limit for constraint force & m\_upperLimit \\ \hline
\end {tabular}}
\label{tab:constraintParameters} 
\end {table}

In structural analysis, a formulation and associated numerical solution procedure are selected 
based on needed features.
Often,  finite element method is used.
In most cases, a static solution with an assumption of linear strain-displacement relation
using displacement based boundary conditions is used.
\cite{bathe-1975} provides a description for handling of various nonlinearities.
In large displacement analysis, formulation may be based on updated formulation (Eulerian) or
Lagrangian formulation where initial configuration is used.
Further enhancements are material nonlinearity and dynamic analysis.
Physics engine provides dynamic analysis with large reference translations and rotations
while assuming bodies to be undeformable.

Material plasticity can be accounted for in simulations by using a suitable coefficient of restitution.
This provides a reasonable means to simulate loss of energy in collisions.
In this work simulation of breaking of bodies made of ductile material is made more realistic 
by splitting the rigid body
to multiple bodies that are connected by energy absorbing joints.
A typical engineering stress-strain curve of ductile steel is shown in Figure \ref{fig:sscurve}.

@@

In Figure \ref{fig:sscurve}, $\sigma$ is stress, $E$ is Youngs modulus and $f_y$ is yield stress.
Engineering stress and strain mean that original dimensions are used in stress calculation,
\cite{dowling}.
%\citet[p.~108]{dowling}.
The stress-strain curve is not drawn to scale as elastic strain could not be seen as it is typically 
0.001 to 0.005 and fracture strain can be 100 times larger.

In this work, an elastic-fully plastic material model is used in most scenarios.
Having elastic part allows elastic displacements for slender structures. 
Elastic material behavior is ignored in approach introduced in this work if
the deformation is related to ahigher frequency
than integration stability would allow.
It should be noted that geometry
of bodies is not updated during analysis and thus engineering stress-strain properties are used.

@@

As shown in Figure 2.4, if stress is below yield limit $f_y$, stress and strain are linear within the material.
If cross section is fully plastic, stress is assumed to be at yield level over the whole cross section such that 
the plastic section modulus is higher than the elastic section modulus.

In this work, plasticity is handled by defining maximum forces
in Equations \ref{eq:lambdaLow} and  
\ref{eq:lambdaHigh} using plastic capacities, which are defined below.

@@

\begin{figure}
\centering
\lstset{language=octave}
\begin{lstlisting}
b=0.01; h=0.01; fy=200e6;
wpy=fy/2*dblquad(@(x,z)...
 sqrt(x.*x+z.*z),-b/2,b/2,-h/2,h/2)
38.2
\end{lstlisting}

\caption{Calculation of maximum moment  around $\vec{r}_{anc}^{\,i} $ using Octave.}
\label{fig:octave-mp}
\end{figure}

The basic idea introduced in this study can be tested with any framework having motors and hinge constraints.
This can be done by setting the target velocity of the motor to zero and limiting 
the maximum motor impulse to plastic moment multiplied by a timestep.

Further enhancements were created and tested by forking \cbullet\ source code
and adding new constraints, \cite{pbullet}.
Instructions for using  windows executable and  source code are available, \cite{bp}.

>>>>

% Simo.Nikula@gmail.com
\def\mytitle{Breakable Elastic-plastic Constraint for Rigid Body Simulation}
\def\lut{Lappeenranta University of Technology}
\def\

\documentclass{jcgt}

\setciteauthor{Simo Nikula, Aki Mikkola and Timo Björk}
\setcitetitle{\\mytitle}

% Mark submissions with the date of submission using the following line:
\submitted{\today}

% Once an article is accepted accepted, switch to the following line and comment the preceding one. The editor will supply the argument values.
%\accepted{2014-02-07}{2014-02-07}{2014-02-07}{Editor Name}{3}{1}{1}{1}{2014}
%\seturl{http://jcgt.org/published/0003/01/01/}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\newcommand{\cbullet}[0]{Bullet Physics}
\title{\\mytitle}

\author
       {SIMO NIKULA {\upshape and} AKI MIKKOLA {\upshape and} TIMO BJÖRK\\\lut
       }

\teaser{
  \includegraphics[width=2in]{figs/demolisher-done.PNG}
  \label{fig:teaser}
  \caption{Adding ductile breakable constraint to popular open source physics engines \cbullet\ and Box2D
allows more realistic simulation of many scenarios.
}
}

\maketitle
\thispagestyle{firstpagestyle}

\begin{abstract}
\small
This paper introduces simple and efficient method to simulate ductile fracture in existing 
impulse-based physics engines.
Method is based on technique of splitting bodies to multiple pieces and joining them with constraints.
Constraints are described using body dimensions and material parameters.

Sample programs with source code are made available to allow developers already using \cbullet\
and Box2D to add plasticity into their simulations.
\end{abstract}

%-------------------------------------------------------------------------
\section{Introduction}
\label{sec:introduction}

Theory for handling of plasticity in computer graphics has been presented already 1988, \cite{cg1988}. 
Recent paper by Jones  provides extensive listing of related work during past decades, \cite{Jones:2016:EPD}.

In impulse-based phycis engines most breakable scenarios  are implemented by destructing various bodies based on collision
or impulse exceeding predefined limit.
Nevertheless, breaking of steel or reinforced concrete structures using this approach 
is not appropriate if the simulation is to look realistic.

More realistic simulation of ductile destructable bodies is possible but in many cases it would require selecting of new physics engine.

\paragraph{Our approach}
This study will introduce an approach to account for plastic deformation in impulse-based phycis engines.   
Presented methodology does not require significant software development efforts from
game vendors and is thus easily adoptable. 
In the introduced method, plastic deformation takes place if the force or moment exceeds a predefined 
limit, deformation absorbs energy and joint breaks if plastic capacity is exceeded. 
Maximum forces and moments are estimated based on the plastic section modulus.
Joint breaking is based on summing plastic deformation and comparing it to a
predefined material based limit. The elastic part of deformation is modelled by employing 
modification of an existing constraints.

\paragraph{Limitations}
Forces are calculated by dividing impulse by timestep and maximum impulse is calculated
by multiplying maximum force by timestep. 
This means that simulation of quick interactions requires small timestep.
This limitation is not introduced by this approach but is build in feature of impulse-based simulation.

Multiaxial stress is not taken into account.

\section{Description of plasticity in the framework of physics engines}

In this work simulation of breaking of bodies made of ductile material is made more realistic 
by splitting the rigid body
to multiple bodies that are connected by energy absorbing joints.

\paragraph{Stress-strain behaviour of ductile steel}
A typical engineering stress-strain curve of ductile steel is shown in Figure \ref{fig:sscurve}.

@@

In Figure \ref{fig:sscurve}, $\sigma$ is stress, $E$ is Youngs modulus and $f_y$ is yield stress.
Engineering stress and strain mean that original dimensions are used in stress calculation,
\cite{dowling}.
%\citet[p.~108]{dowling}.
The stress-strain curve is not drawn to scale as elastic strain could not be seen as it is typically 
0.001 to 0.005 and fracture strain can be 100 times larger.
In practice this means that elastic displacement due to axial forces is usually not visible.
Visible elastic displacement is usually due to bending.

In this work, an elastic-fully plastic material model is used in most scenarios.
Having elastic part allows elastic displacements for slender structures. 
Elastic material behavior is ignored in approach introduced in this work if
the deformation is related to a higher frequency
than integration stability would allow.
It should be noted that geometry
of bodies is not updated during analysis and thus engineering stress-strain properties are used.

@@

As shown in Figure \ref{fig:wp}, if stress is below yield limit $f_y$, stress and strain are linear within the material.
If cross section is fully plastic, stress is assumed to be at yield level over the whole cross section such that 
the plastic section modulus is higher than the elastic section modulus.

\paragraph{Plastic capacities}
In this work, plasticity is handled by defining maximum forces using plastic capacities, which are defined below.

@@

\lstset{language=octave}
\begin{lstlisting}[caption={Calculation of maximum moment  around $\vec{r}_{anc}^{\,i} $ using Octave.}, 
label={lst:octave-mp}, float]
b=0.01; h=0.01; fy=200e6;
wpy=fy/2*dblquad(@(x,z)...
 sqrt(x.*x+z.*z),-b/2,b/2,-h/2,h/2)
38.2
\end{lstlisting}

\paragraph{Joint breaking}

\section{Implementation for Box2D}

\section{Implementation for \cbullet\}

\small
\bibliographystyle{jcgt}
\bibliography{paper}

\section*{Index of Supplemental Materials}

\section*{Author Contact Information}

\hspace{-2mm}\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
Simo Nikula\newline
\href{mailto:Simo.Nikula@gmail.com}{Simo.Nikula@gmail.com}
&
Aki Mikkola\newline
\href{mailto:Aki.Mikkola@lut.fi}{Aki.Mikkola@lut.fi}
&
Timo Björk\newline
\href{mailto:Timo.Bjork@lut.fi}{Timo.Bjork@lut.fi}

\end{tabular}

\afterdoc

\end{document}
